Do investors in clean energy ETFs herd?

### 2025-01-07 alt_energy_herding.qmd
complete draft

In this new climate sustainability paradigm, investors face many pressures that not only bear on returns but also the stability of financial markets. For example, resulting regulations aimed at reducing emissions can surprisingly reduce the profitability of fossil-fuel-based companies, or the possible mispricing of assets from ignoring climate risks can lead to significant losses [@nguyen2025]. In addition to these climate risks, a general change in investor attitudes can drive the inclusion of green assets in their portfolio leading to systemic risk.^[These pressures notwithstanding the possible contribution that financial markets can play in mitigating and reducing the negative effects of climate change [@giglio2021a].

@@

Exchange-traded funds (ETFs) are a key feature of green financial products. ETFs are a type of security that involves a collection of securities that often tracks an underlying index. However, they can invest in various industry sectors or strategies. In addition, environment, social, and governance (ESG) ETFs serve as a market discovery tool for investors to identify and invest in environmentally friendly companies [@briere2023]. Among the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing ones in 2022, followed by the Cybersecurity and Artificial Intelligence (AI) ETFs [@decclesia2024]. The clean energy transition represents one of the largest multi-decade secular growth opportunities.  After the inclusion of Green energy financing in the list of United Nations Sustainability Goals (SDGs) as SDG 7, the role, importance, and visibility of green financial products; all have escalated enormously [@naqvi2022]. Therefore, the growth of green assets under management is likely to continue.

However, given how recent the inclusion of climate sustainability investment decisions is, it is not clear what the actual impact will be in the long run. In this study specifically, we ask whether the rapid adoption of CE ETFs could be driven by market fads, or is a fundamental change in investor behaviour. Investors, for example, can believe that peers have information about climate risks that they do not, investors may herd to avoid losses compared to peers, or investors may be encouraged to herd by the desire to align to climate-related social values [@ciciretti2021; @gavrilakis2023; @loang2023]. Therefore, market volatility and crisis, financial performance, and investor sentiment can drive herding in ESG markets.

@@

However, to the best of our knowledge, no studies exist that focus on herding behaviour in CE ESGs. This study, therefore, aims to extend the broader literature on herding behaviour in commodity and energy markets. Several studies in this area were conducted. For example, @demirer2013a conducted a sector study and found herding behaviour in grains but not in other sectors. Similarly, @gilbert2010 showed herding behaviour amongst speculators in non-ferrous commodities. Others did not find evidence of herding in similar markets. @babalos2015 found significant anti-herding behaviour in metal commodities futures after the global financial crisis. @pierdzioch2010 showed that forecasters in oil and metals markets deviated from the crowd, indicating a rational response to information. @steen2013 also found no herding behaviour in international commodity markets. Overall, the literature in this area is mixed, which indicates scope for further study.

@@

Stated differently, as @chang2000 argue, in the case of herding the coefficient on the non-linear term ($\gamma_2$) will be negative and statistically significant. @tbl-herding presents the results of herding for the full sample employing the non-linear @eq-herding. The estimated coefficient on market return is positive and highly significant as expected.The estimated coefficient on the non-linear term is negative (-1.2773) and statistically significant with a t-statistic of -9.71 suggesting that herd behaviour is present and robust in the US alternative energy ETFs.

@@

There is ample evidence in the relevant literature that herding behaviour in various asset markets [see @pochea2017] exhibits asymmetry and time-varying characteristics. To this end, we proceed to estimate @eq-herding using the quantile regression (QR) proposed by @koenker1978 and @tbl-quantile_herding presents the results of estimating @eq-herding across various quantiles of the returns dispersion. Our focus is on the herding coefficient $\gamma_2$, as a significant negative value of $\gamma_2$ is indicative of herding. Such a finding is observed at two quantiles namely 25% and 50% with a value of -1.1056 and -1.165 which are highly significant. It is worth mentioning that the sign of the herding coefficient remains negative for almost all quantiles while the significance changes from significant to insignificant while we move from low and middle to upper quantiles (75\% and 90\%).

@@

Herding is present at low quantiles when markets are rising with an estimated coefficient $\gamma_3$ of -1.3669 and -1.2383 and highly significant respectively. However, when markets are declining, investors seem to neglect their own information set and imitate the actions of others resulting in a highly significant coefficient of herding ($\gamma_4$) across four out of five quantiles. Furthermore, we find that in high quantiles (75\% and 90\%) and when markets are rising the coefficient of interest ($\gamma_3$) turns positive but insignificant.

@@

We use these high PRI and high TRI in one probit regression and low TRI and low PRI in another. Results are presented @tbl-probit_analysis2. We observe higher uncertainty that stems from physical or transition risk causes anti-herding which is in line with the logic we discussed earlier.

@@

This study offers novel and valuable insights on herding behavior in clean energy ETFs.

Results of baseline and rolling window analysis points to significant herding for the whole period and in various instances.

@choi2023
@kuang2021

>>>>

In this new climate sustainability paradigm, investors face many pressures that not only bear on returns but also the stability of financial markets. For example, resulting regulations aimed at reducing emissions can surprisingly reduce the profitability of fossil-fuel-based companies, or the possible mispricing of assets from ignoring climate risks can lead to significant losses [@nguyen2025]. In addition to these climate risks, a general change in investor attitudes can drive the inclusion of green assets in their portfolio can lead to systemic risk.^[These pressures notwithstanding the possible contribution that financial markets can play in mitigating and reducing the negative effects of climate change [@giglio2021a]]

@@

Exchange-traded funds (ETFs) are a key feature of green financial products. ETFs are a type of security that involves a collection of securities that often tracks an underlying index. However, they can invest in various industry sectors or strategies. In addition, environment, social, and governance (ESG) ETFs serve as a market discovery tool for investors to identify and invest in environmentally friendly companies [@briere2023]. Among the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing ones in 2022, followed by the Cybersecurity and Artificial Intelligence (AI) ETFs [@decclesia2024]. The clean energy transition represents one of the largest multi-decade secular growth opportunities.  After the inclusion of Green energy financing in the list of United Nations Sustainability Goals (SDGs) as SDG 7, the role, importance, and visibility of green financial products have escalated enormously [@naqvi2022]. That is, the growth of green assets under management is likely to continue.

However, given how recent the inclusion of climate sustainability in investment decisions is, it is not clear what the actual impact will be in the long run. In this study specifically, we ask whether the rapid adoption of CE ETFs could be driven by market fads, or is a fundamental change in investor behaviour. Investors, for example, can believe that peers have information about climate risks that they do not, investors may herd to avoid losses compared to peers, or investors may be encouraged to herd by the desire to align to climate-related social values [@ciciretti2021; @gavrilakis2023; @loang2023]. Therefore, market volatility and crisis, financial performance, and investor sentiment can drive herding in ESG markets.

@@

However, to the best of our knowledge, no studies exist that focus on herding behaviour in CE ESGs. This study, therefore, aims to extend the broader literature on herding behaviour in commodity and energy markets. Several studies in this area were conducted. For example, @demirer2013a conducted a commodity sectoral study and found herding behaviour in grains but not in other sectors. Similarly, @gilbert2010 showed herding behaviour amongst speculators in non-ferrous commodities. Others did not find evidence of herding in similar markets. @babalos2015 found significant anti-herding behaviour in metal commodities futures after the global financial crisis. @pierdzioch2010 showed that forecasters in oil and metals markets deviated from the crowd, indicating a rational response to market information. @steen2013 also found no herding behaviour in international commodity markets. Overall, the literature in this area is mixed, which indicates scope for further study.

@@

Stated differently, as @chang2000 argue, in the case of herding the coefficient on the non-linear term ($\gamma_2$) will be negative and statistically significant. @tbl-herding presents the results of herding for the full sample employing the non-linear @eq-herding. The estimated coefficient on market return is positive and highly significant as expected. The estimated coefficient on the non-linear term is negative (-1.2773) and statistically significant with a t-statistic of -9.71 suggesting that herd behaviour is present and robust in the US alternative energy ETFs.

@@

There is ample evidence in the relevant literature that herding behaviour in various asset markets [see @pochea2017] exhibits asymmetry and time-varying characteristics. To this end, we proceed to estimate @eq-herding using the quantile regression (QR) proposed by @koenker1978. @tbl-quantile_herding presents the results of estimating @eq-herding across various quantiles of the returns dispersion. Our focus is on the herding coefficient $\gamma_2$, as a significant negative value of $\gamma_2$ is indicative of herding. Such a finding is observed at two quantiles namely 25% and 50% with a value of -1.1056 and -1.165 which are highly significant. It is worth mentioning that the sign of the herding coefficient remains negative for almost all quantiles while the significance changes from significant to insignificant while we move from low and middle to upper quantiles (75\% and 90\%).

@@

Herding is present at all quantiles when markets are rising with an estimated coefficient $\gamma_3$ of -1.3669 and -1.2383 and highly significant respectively. However, when markets are declining, investors seem to neglect their own information set and imitate the actions of others resulting in a highly significant coefficient of herding ($\gamma_4$) across four out of five quantiles. Furthermore, we find that in high quantiles (75\% and 90\%) and when markets are rising the coefficient of interest ($\gamma_3$) turns positive but insignificant.

@@

We use these high PRI and high TRI in one probit regression and low TRI and low PRI in another. Results are presented in @tbl-probit_analysis2. We observe higher uncertainty that stems from physical or transition risk causes anti-herding which is in line with the logic we discussed earlier.

@@

This study offers novel and valuable insights into herding behaviour in clean energy ETFs. We used various herding behaviour tests to achieve this. First, the baseline herding tests revealed significant evidence of herding behaviour. Second, the asymmetric herding tests showed that herding behaviour was present in both bear and bull markets. Lastly, the results of the time-varying tests point to significant periodic herding. However, the results further indicate that herding in US clean energy ETFs was not mainly related to climate-related risks or uncertainty. The probit analysis showed that climate-related uncertainty reduced the probability of herding or led to investor anti-herding behaviour. We defer to the literature on the factors that drive herding in these ETFs [for example, @loang2023;  @koutmos2024; @przychodzen2016]. Our results, similar to @bua2024, indicate that investors did not herd on climate-related uncertainty, and sought returns whilst hedging against climate-related risk. This is a positive for the financial stability of these growing alternative investment indices.

### 2025-01-07 alt_energy_herding/alt_energy_herding.tex
complete draft

\author { 
Vassilios Babalos\footnote{Corresponding Author. Department of Accounting and Finance, University of Peloponnese, Kalamata,  Greece; Email: v.babalos@uop.gr.} \,\,
Xolani Sibande\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: xolaniss@gmail.com. Economic Research Department, South African Reserve Bank, Pretoria, South Africa.} \,\, 
Elie Bouri\footnote{Adnan Kassar School of Business, Lebanese American University, Lebanon; Email: elie.elbouri@lau.edu.lb} \,\,
Rangan Gupta\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: rangan.gupta@up.ac.za.} 
}
\date{\today}
\maketitle

@@

ipsum

@@

\noindent\textbf{Keywords}: Herding Behaviour, Financial Crises, Market Efficiency
\\
\textbf{JEL Codes}: G01, G14, G15, G41
\newpage

@@

In this new climate sustainability paradigm, investors face many
pressures that not only bear on returns but also the stability of
financial markets. For example, resulting regulations aimed at reducing
emissions can surprisingly reduce the profitability of fossil-fuel-based
companies, or the possible mispricing of assets from ignoring climate
risks can lead to significant losses \citep{nguyen2025}. In addition to
these climate risks, a general change in investor attitudes can drive
the inclusion of green assets in their portfolio leading to systemic
risk.\^{}{[}These pressures notwithstanding the possible contribution
that financial markets can play in mitigating and reducing the negative
effects of climate change \citep{giglio2021a}.

@@

Exchange-traded funds (ETFs) are a key feature of green financial
products. ETFs are a type of security that involves a collection of
securities that often tracks an underlying index. However, they can
invest in various industry sectors or strategies. In addition,
environment, social, and governance (ESG) ETFs serve as a market
discovery tool for investors to identify and invest in environmentally
friendly companies \citep{briere2023}. Among the ESG ETFs, the Clean
Energy (CE) ETFs have been the best-performing ones in 2022, followed by
the Cybersecurity and Artificial Intelligence (AI) ETFs
\citep{decclesia2024}. The clean energy transition represents one of the
largest multi-decade secular growth opportunities. After the inclusion
of Green energy financing in the list of United Nations Sustainability
Goals (SDGs) as SDG 7, the role, importance, and visibility of green
financial products; all have escalated enormously \citep{naqvi2022}.
Therefore, the growth of green assets under management is likely to
continue.

However, given how recent the inclusion of climate sustainability
investment decisions is, it is not clear what the actual impact will be
in the long run. In this study specifically, we ask whether the rapid
adoption of CE ETFs could be driven by market fads, or is a fundamental
change in investor behaviour. Investors, for example, can believe that
peers have information about climate risks that they do not, investors
may herd to avoid losses compared to peers, or investors may be
encouraged to herd by the desire to align to climate-related social
values \citep{ciciretti2021, gavrilakis2023, loang2023}. Therefore,
market volatility and crisis, financial performance, and investor
sentiment can drive herding in ESG markets.

@@

However, to the best of our knowledge, no studies exist that focus on
herding behaviour in CE ESGs. This study, therefore, aims to extend the
broader literature on herding behaviour in commodity and energy markets.
Several studies in this area were conducted. For example,
\citet{demirer2013a} conducted a sector study and found herding
behaviour in grains but not in other sectors. Similarly,
\citet{gilbert2010} showed herding behaviour amongst speculators in
non-ferrous commodities. Others did not find evidence of herding in
similar markets. \citet{babalos2015} found significant anti-herding
behaviour in metal commodities futures after the global financial
crisis. \citet{pierdzioch2010} showed that forecasters in oil and metals
markets deviated from the crowd, indicating a rational response to
information. \citet{steen2013} also found no herding behaviour in
international commodity markets. Overall, the literature in this area is
mixed, which indicates scope for further study.

@@

Stated differently, as \citet{chang2000} argue, in the case of herding
the coefficient on the non-linear term (\(\gamma_2\)) will be negative
and statistically significant. Table~\ref{tbl-herding} presents the
results of herding for the full sample employing the non-linear
Equation~\ref{eq-herding}. The estimated coefficient on market return is
positive and highly significant as expected.The estimated coefficient on
the non-linear term is negative (-1.2773) and statistically significant
with a t-statistic of -9.71 suggesting that herd behaviour is present
and robust in the US alternative energy ETFs.

@@

There is ample evidence in the relevant literature that herding
behaviour in various asset markets \citep[see][]{pochea2017} exhibits
asymmetry and time-varying characteristics. To this end, we proceed to
estimate Equation~\ref{eq-herding} using the quantile regression (QR)
proposed by \citet{koenker1978} and Table~\ref{tbl-quantile_herding}
presents the results of estimating Equation~\ref{eq-herding} across
various quantiles of the returns dispersion. Our focus is on the herding
coefficient \(\gamma_2\), as a significant negative value of
\(\gamma_2\) is indicative of herding. Such a finding is observed at two
quantiles namely 25\% and 50\% with a value of -1.1056 and -1.165 which
are highly significant. It is worth mentioning that the sign of the
herding coefficient remains negative for almost all quantiles while the
significance changes from significant to insignificant while we move
from low and middle to upper quantiles (75\% and 90\%).

@@

Herding is present at low quantiles when markets are rising with an
estimated coefficient \(\gamma_3\) of -1.3669 and -1.2383 and highly
significant respectively. However, when markets are declining, investors
seem to neglect their own information set and imitate the actions of
others resulting in a highly significant coefficient of herding
(\(\gamma_4\)) across four out of five quantiles. Furthermore, we find
that in high quantiles (75\% and 90\%) and when markets are rising the
coefficient of interest (\(\gamma_3\)) turns positive but insignificant.

@@

We use these high PRI and high TRI in one probit regression and low TRI
and low PRI in another. Results are presented
Table~\ref{tbl-probit_analysis2}. We observe higher uncertainty that
stems from physical or transition risk causes anti-herding which is in
line with the logic we discussed earlier.

@@

This study offers novel and valuable insights on herding behavior in
clean energy ETFs.

Results of baseline and rolling window analysis points to significant
herding for the whole period and in various instances.

\citet{choi2023} \citet{kuang2021}

>>>>

\author{ 
Vassilios Babalos\footnote{Corresponding Author. Department of Accounting and Finance, University of Peloponnese, Kalamata,  Greece; Email: v.babalos@uop.gr.} \,\,
Xolani Sibande\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: xolaniss@gmail.com. Economic Research Department, South African Reserve Bank, Pretoria, South Africa.} \,\, 
Elie Bouri\footnote{Adnan Kassar School of Business, Lebanese American University, Lebanon; Email: elie.elbouri@lau.edu.lb} \,\,
Rangan Gupta\footnote{Department of Economics, University of Pretoria, Pretoria, South Africa; Email: rangan.gupta@up.ac.za.} 
}
\date{\today}
\maketitle

@@

This study offers novel and valuable insights into herding behaviour in US clean energy ETFs between 2016 and 2023. 
The baseline herding tests by \cite{christie1995} and \cite{chang2000} revealed significant herding behaviour in this market. 
This evidence was supported by asymmetric and time-varying herding tests. That is, investors herd both in bear and bullish markets, and periodically. 
In addition, we found that climate risks (both physical and transition) reduced the probability of herding in US clean energy ETFs, 
indicating that an increase in climate-related risk encouraged efficient or climate-hedging behaviour by investors. 
Therefore, the results suggest that climate-related uncertainty did not drive herding behaviour in this market. The results suggest that investors are appropriately identifying opportunities that mitigate climate risk,
thereby reducing the probability of herding-driven system risk.

@@

\noindent\textbf{Keywords}: Herding Behaviour, Climate Change, Energy
\\
\textbf{JEL Codes}: G14, Q54, P18
\newpage

@@

In this new climate sustainability paradigm, investors face many
pressures that not only bear on returns but also the stability of
financial markets. For example, resulting regulations aimed at reducing
emissions can surprisingly reduce the profitability of fossil-fuel-based
companies, or the possible mispricing of assets from ignoring climate
risks can lead to significant losses \citep{nguyen2025}. In addition to
these climate risks, a general change in investor attitudes can drive
the inclusion of green assets in their portfolio can lead to systemic
risk.\footnote{These pressures notwithstanding the possible contribution
  that financial markets can play in mitigating and reducing the
  negative effects of climate change \citep{giglio2021a}}

@@

Exchange-traded funds (ETFs) are a key feature of green financial
products. ETFs are a type of security that involves a collection of
securities that often tracks an underlying index. However, they can
invest in various industry sectors or strategies. In addition,
environment, social, and governance (ESG) ETFs serve as a market
discovery tool for investors to identify and invest in environmentally
friendly companies \citep{briere2023}. Among the ESG ETFs, the Clean
Energy (CE) ETFs have been the best-performing ones in 2022, followed by
the Cybersecurity and Artificial Intelligence (AI) ETFs
\citep{decclesia2024}. The clean energy transition represents one of the
largest multi-decade secular growth opportunities. After the inclusion
of Green energy financing in the list of United Nations Sustainability
Goals (SDGs) as SDG 7, the role, importance, and visibility of green
financial products have escalated enormously \citep{naqvi2022}. That is,
the growth of green assets under management is likely to continue.

However, given how recent the inclusion of climate sustainability in
investment decisions is, it is not clear what the actual impact will be
in the long run. In this study specifically, we ask whether the rapid
adoption of CE ETFs could be driven by market fads, or is a fundamental
change in investor behaviour. Investors, for example, can believe that
peers have information about climate risks that they do not, investors
may herd to avoid losses compared to peers, or investors may be
encouraged to herd by the desire to align to climate-related social
values \citep{ciciretti2021, gavrilakis2023, loang2023}. Therefore,
market volatility and crisis, financial performance, and investor
sentiment can drive herding in ESG markets.

@@

However, to the best of our knowledge, no studies exist that focus on
herding behaviour in CE ESGs. This study, therefore, aims to extend the
broader literature on herding behaviour in commodity and energy markets.
Several studies in this area were conducted. For example,
\citet{demirer2013a} conducted a commodity sectoral study and found
herding behaviour in grains but not in other sectors. Similarly,
\citet{gilbert2010} showed herding behaviour amongst speculators in
non-ferrous commodities. Others did not find evidence of herding in
similar markets. \citet{babalos2015} found significant anti-herding
behaviour in metal commodities futures after the global financial
crisis. \citet{pierdzioch2010} showed that forecasters in oil and metals
markets deviated from the crowd, indicating a rational response to
market information. \citet{steen2013} also found no herding behaviour in
international commodity markets. Overall, the literature in this area is
mixed, which indicates scope for further study.

@@

Stated differently, as \citet{chang2000} argue, in the case of herding
the coefficient on the non-linear term (\(\gamma_2\)) will be negative
and statistically significant. Table~\ref{tbl-herding} presents the
results of herding for the full sample employing the non-linear
Equation~\ref{eq-herding}. The estimated coefficient on market return is
positive and highly significant as expected. The estimated coefficient
on the non-linear term is negative (-1.2773) and statistically
significant with a t-statistic of -9.71 suggesting that herd behaviour
is present and robust in the US alternative energy ETFs.

@@

There is ample evidence in the relevant literature that herding
behaviour in various asset markets \citep[see][]{pochea2017} exhibits
asymmetry and time-varying characteristics. To this end, we proceed to
estimate Equation~\ref{eq-herding} using the quantile regression (QR)
proposed by \citet{koenker1978}. Table~\ref{tbl-quantile_herding}
presents the results of estimating Equation~\ref{eq-herding} across
various quantiles of the returns dispersion. Our focus is on the herding
coefficient \(\gamma_2\), as a significant negative value of
\(\gamma_2\) is indicative of herding. Such a finding is observed at two
quantiles namely 25\% and 50\% with a value of -1.1056 and -1.165 which
are highly significant. It is worth mentioning that the sign of the
herding coefficient remains negative for almost all quantiles while the
significance changes from significant to insignificant while we move
from low and middle to upper quantiles (75\% and 90\%).

@@

Herding is present at all quantiles when markets are rising with an
estimated coefficient \(\gamma_3\) of -1.3669 and -1.2383 and highly
significant respectively. However, when markets are declining, investors
seem to neglect their own information set and imitate the actions of
others resulting in a highly significant coefficient of herding
(\(\gamma_4\)) across four out of five quantiles. Furthermore, we find
that in high quantiles (75\% and 90\%) and when markets are rising the
coefficient of interest (\(\gamma_3\)) turns positive but insignificant.

@@

We use these high PRI and high TRI in one probit regression and low TRI
and low PRI in another. Results are presented in
Table~\ref{tbl-probit_analysis2}. We observe higher uncertainty that
stems from physical or transition risk causes anti-herding which is in
line with the logic we discussed earlier.

@@

This study offers novel and valuable insights into herding behaviour in
clean energy ETFs. We used various herding behaviour tests to achieve
this. First, the baseline herding tests revealed significant evidence of
herding behaviour. Second, the asymmetric herding tests showed that
herding behaviour was present in both bear and bull markets. Lastly, the
results of the time-varying tests point to significant periodic herding.
However, the results further indicate that herding in US clean energy
ETFs was not mainly related to climate-related risks or uncertainty. The
probit analysis showed that climate-related uncertainty reduced the
probability of herding or led to investor anti-herding behaviour. We
defer to the literature on the factors that drive herding in these ETFs
\citep[for example,][]{loang2023, koutmos2024, przychodzen2016}. Our
results, similar to \citet{bua2024}, indicate that investors did not
herd on climate-related uncertainty, and sought returns whilst hedging
against climate-related risk. This is a positive for the financial
stability of these growing alternative investment indices.
